ANCOVA Comparison Simulations: Nonlinear Model

### 2017-10-05 simulations/nonlinear_model.Rmd
Nonlinear model simulations

---
title: "ANCOVA Comparison Simulations: Discrete Outcomes"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

# Discrete data

Step 1:
We generated the covariate sand potential outcomes according to the methods in the continuous outcome section but discretized them by taking the floor.
Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ 10,000 times.
We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

Step 2:
We regenerate $Z$ and recompute $Y_(Z)$ 10,000 times for each design.

In expectation, the average treatment effect is $\gamma$.
We compare the empirical power of five tests to detect this treatment effect:

* ANCOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
* Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and stratum dummies
* Freedman-Lane test: see the other Rmd document for a full description of this procedure

@@

```{r generate_sample}
gen_y <- function(gamma, v, error, Z){
  y <- floor(0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error)
  return(y)
}

gen_x <- function(gamma, v, error){
  x <- floor(0.5*(-1*gamma*exp(v) + exp(v/2)) + error)
  return(x)
}

@@

  # Generate covariates
  Z <- rep(0:1, length.out = N)
  Y0 <- gen_x(gamma, v, epsilon)
  Y1 <- gen_y(gamma, v, delta, Z)
  return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
}

@@

# Constant additive treatment effect

```{r vary_effect_size_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
```

First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
We varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
In the random populations that were generated, this corresponded to population average treatment effects of `r round(tr_effect, 2)[1:4]`, and `r round(tr_effect, 2)[5]`, respectively.
Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
When the effect size was zero, all five tests had the correct level of 5\%, though the stratified and differenced permutation tests rejected 3.5\% of the time.
The differenced permutation test had substantially lower power than the rest,
again because the correlation between baseline $X$ and outcome $Y$ was low.
In this case, the unadjusted stratified test had comparable power to the linear model based tests.

# Heterogeneous treatment effect

```{r vary_effect_size_het_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_het_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
```
Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
The pattern here is nearly identical to Figure 2 in the continuous outcome simulations, so we leave out further comments.

```{r plot_for_ms, include=FALSE}
p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
  ylim(0, 0.95 ) +
  theme(legend.position="bottom") + 
  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  report_theme +
  ggtitle("Constant Treatment \n Effect")

p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
    ylim(0, 0.95) +
    report_theme +
    ggtitle("Heterogeneous Treatment \n Effect")

tmp <- ggplot_gtable(ggplot_build(p1))
leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
mylegend <- tmp$grobs[[leg]]

p3 <- grid.arrange(arrangeGrob(p1 + theme(legend.position="none"),
                         p2 + theme(legend.position="none"),
                         nrow=1),
             mylegend, nrow=2,heights=c(10, 2))

p3
ggsave("../ms/fig/discrete_outcomes_simulation_power.pdf", p3, device="pdf", width=8)

>>>>

---
title: "ANCOVA Comparison Simulations: Nonlinear Model"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

The potential outcomes model we used before turned out to be a linear function of $X$ and $Z$, conditional on the latent variable $v$.
For these simulations, we changed the outcome model to be nonlinear.
Instead of an additive treatment effect, now the effect is multiplicative.
We generated the latent variable and errors in the same way, but let

\begin{align*}
X_{ij} &=  \frac{ e^{v_{ij}} + e^{v_{ij}/2}}{2} +\varepsilon_{ij} \\
Y_{ij}(Z_{ij}) &= (1+\gamma)^{Z_{ij}} X_{ij}  + \delta{ij}.
\end{align*}

The treatment effect for individual $(i, j)$ is thus $\gamma X_{ij}$.

@@

```{r generate_sample}

gen_y <- function(gamma, x, error, Z){
  multiplier <- (1+gamma)^Z
  y <- multiplier*x + error
  return(y)
}

gen_x <- function(gamma, v, error){
  x <- 0.5*(exp(v) + exp(v/2)) + error
  return(x)
}

@@

  # Generate covariates
  Z <- rep(0:1, length.out = N)
  Y0 <- gen_y0(gamma, v, epsilon)
  Y1 <- gen_y1(gamma, v, delta, Z)
  return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
}

@@

# Homogeneous treatment effects

```{r vary_effect_size_plot, fig.height=4}
load("nonlinear_model/vary_gamma_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
```

First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
We varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
In the random populations that were generated, this corresponded to population average treatment effects of `r round(tr_effect, 2)[1:4]`, and `r round(tr_effect, 2)[5]`, respectively.
Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
When the effect size was zero, all five tests had the correct level of 5\%.
The stratified, unadjusted permutation test failed here:
the treatment effect is entangled with the baseline covariate, so failing to account for the baseline obscures the treatment effect.
The parametric and permutation linear model tests perform comparably and are plotted on top of each other, while the differenced permutation test has slightly more power.

# Heterogeneous treatment effect
```{r vary_effect_size_het_plot, fig.height=4}
load("nonlinear_model/vary_gamma_het_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
```
Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
Four of the five tests had the correct level of 5\%, while the stratified permutation test rejected only 1\% of the time.
The same pattern appeared here: the stratified permutation test rarely rejects, while the differenced permutation test dominates all of the linear model based tests.
